\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}

\title{Surgical Projection: Mathematical Formulation}
\author{}
\date{}

\begin{document}
\maketitle

\section{Original (Full-Alias) Method}

Given a writer matrix $\mathbf{W}$ ($\mathbf{W}_O$ or $\mathbf{W}_{\text{down}}$), its mean calibration input $\mathbf{x}$, and a random alias $\mathbf{z} \in \mathbb{R}^d$:

\begin{equation}
    \text{target} = \mathbf{z}, \quad \mathbf{z} \sim \mathcal{N}(0, \varepsilon^2 I)
\end{equation}

The rank-one update makes $\mathbf{W}_{\text{new}} \mathbf{x} = \mathbf{z}$. The pollution added to the residual stream is:

\begin{equation}
    \boldsymbol{\Delta} = \mathbf{z} - \mathbf{W}\mathbf{x}
\end{equation}

This is a full $d$-dimensional vector. It is large because $\mathbf{z}$ has no relation to the original output---the entire $d$-dimensional output is replaced with random noise.

\section{Surgical Projection Method}

Decompose the writer's current output into its component along the refusal direction $\hat{r}$ and everything orthogonal:

\begin{equation}
    \mathbf{W}\mathbf{x} = \underbrace{(\mathbf{W}\mathbf{x} \cdot \hat{r})\,\hat{r}}_{\text{refusal component}} + \underbrace{\mathbf{W}\mathbf{x} - (\mathbf{W}\mathbf{x} \cdot \hat{r})\,\hat{r}}_{\text{everything else}}
\end{equation}

We only replace the scalar projection with a random value $\eta \sim \mathcal{N}(0, \varepsilon^2)$:

\begin{equation}
    \text{target} = \mathbf{W}\mathbf{x} - (\mathbf{W}\mathbf{x} \cdot \hat{r})\,\hat{r} + \eta\,\hat{r}
\end{equation}

The pollution is now:

\begin{equation}
    \boldsymbol{\Delta} = \text{target} - \mathbf{W}\mathbf{x} = (\eta - \mathbf{W}\mathbf{x} \cdot \hat{r})\,\hat{r}
\end{equation}

This is a vector \textbf{purely along} $\hat{r}$ with scalar magnitude $|\eta - \mathbf{W}\mathbf{x} \cdot \hat{r}|$.

\section{Why This Matters}

\begin{enumerate}[leftmargin=*, label=\textbf{\arabic*.}]
    \item \textbf{Pollution is 1-dimensional} instead of $d$-dimensional ($d = 4096$). The cumulative $\mathbf{z}_{\text{sum}}$ stays along $\hat{r}$, so reader patches only need to correct for a tiny perturbation along one direction rather than an arbitrary perturbation across the full space.

    \item \textbf{Off-calibration drift is minimal.} The rank-one update is exact at $\mathbf{x}$ but approximate for other inputs. When the pollution is small ($\|\boldsymbol{\Delta}\| = 0.47$ vs $61.3$), the reader patches' approximation error on non-calibration inputs is proportionally smaller---which is why harmless KL drops from $15.4$ to $0.007$.

    \item \textbf{The refusal signal is still obfuscated.} At the calibration point, each writer's output along $\hat{r}$ becomes $\eta$ (random noise) instead of its original projection. An attacker running difference-in-means on the defended model's writer outputs gets randomized $\hat{r}$ components, so the extracted direction no longer cleanly aligns with the true refusal direction.

    \item \textbf{Non-refusal behavior is preserved exactly} (at the calibration point). The orthogonal component $\mathbf{W}\mathbf{x} - (\mathbf{W}\mathbf{x} \cdot \hat{r})\hat{r}$ is untouched, so the model's non-refusal computations pass through unchanged.
\end{enumerate}

\end{document}
